---
title: "R for Clinical Programming: Comparison, Files, Functions, Git & Packages"
subtitle: "A Detailed, Beginner-Friendly Walkthrough of package_dev.R"
author: "Ram"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This tutorial explains, line by line and in plain language, the ideas in the
`package_dev.R` script. It is written for **beginners** — every function is
introduced with *what it does*, *why you'd use it*, and *a runnable example*.

The examples use a clinical-programming flavour (comparing a "development" data
set against a "QC" data set, reading SDTM `.sas7bdat` files), but the R concepts
apply to any field.

> **How to read this document.** Chunks that create files, read from your disk,
> or talk to Git are marked `eval=FALSE` so the document knits safely. Copy them
> into your own RStudio session to run them for real. The small illustrative
> chunks evaluate live when you knit.

---

# Comparing Two Data Frames

A daily task in clinical programming is **double programming**: two people build
the same data set independently, and you check that the results match. R has
several tools for comparing data frames.

## Set up two sample data frames

```{r compare-setup}
# "development" version
adsl_dev <- data.frame(
  USUBJID = c("01", "02", "03", "04"),
  AGE     = c(34, 45, 29, 50)
)

# "QC" version — note subject 04 has AGE 51 instead of 50
adsl_qc <- data.frame(
  USUBJID = c("01", "02", "03", "04"),
  AGE     = c(34, 45, 29, 51)
)

adsl_dev
adsl_qc
```

The two frames are identical except for one value. Our goal is to have R *tell
us* exactly where they differ.

## The quickest check: `identical()`

`identical()` is base R. It returns a single `TRUE`/`FALSE` — the frames either
match exactly or they don't. It won't tell you *where* they differ, but it's a
fast first gate.

```{r compare-identical}
identical(adsl_dev, adsl_qc)   # FALSE, because AGE differs for subject 04
```

## Detailed differences with `diffdf`

The **diffdf** package is built for exactly this job. `diffdf()` reports which
variables differ, in which rows, and shows the mismatched values side by side.

```{r compare-diffdf, eval=FALSE}
install.packages("diffdf")   # one-time install
library(diffdf)

diffdf(adsl_dev, adsl_qc)
```

Typical output points you straight to the problem:

```
Differences found between the two datasets...
Variable  No of Differences  First Difference
AGE       1                  Row 4
...
VARIABLE: AGE
   ..ROW..   BASE   COMPARE
   4         50     51
```

You can **save the result** to an object to inspect or log it later:

```{r compare-diffdf-save, eval=FALSE}
diff_output <- diffdf(adsl_dev, adsl_qc)
diff_output   # re-print the stored comparison anytime
```

## An alternative view with `arsenal`

The **arsenal** package's `comparedf()` gives a different, summary-style report.
The `by =` argument tells it which column identifies a row (here the subject ID),
so it compares like-for-like subjects rather than relying on row order.

```{r compare-arsenal, eval=FALSE}
install.packages("arsenal")   # one-time install
library(arsenal)

summary(comparedf(adsl_dev, adsl_qc, by = "USUBJID"))
```

`diffdf` vs `arsenal`: both are excellent. `diffdf` is compact and popular in
pharma QC; `arsenal::comparedf()` produces a richer, publication-style summary.
Use whichever your team standardises on.

---

# Talking to the User: message, print, warning, stop

R has four related ways for code to communicate. Understanding the difference is
important because they behave very differently — especially `stop()`.

| Function | Purpose | Stops execution? | Goes to |
|---|---|---|---|
| `print()` | Show a value/result | No | Console (stdout) |
| `message()` | Friendly status/progress note | No | Console (stderr) |
| `warning()` | "Something looks off, but I continued" | No | Collected, shown after |
| `stop()` | Fatal error — halt immediately | **Yes** | Error stream |

```{r messaging, error=TRUE}
message("Comparison of adsl_dev and adsl_qc completed. Differences identified.")
print("Please review the output for details on the discrepancies.")
warning("Ensure the data frames are correctly aligned and differences are expected.")
stop("Please investigate the discrepancies before proceeding.")
```

A few beginner notes:

- `message()` is the polite way to print progress. Because it uses *stderr*, it
  won't get mixed into results you capture, and users can silence it with
  `suppressMessages()`.
- `warning()` does **not** stop your code. R holds warnings and shows them at the
  end; `warnings()` re-displays the most recent ones.
- `stop()` **halts execution immediately** — everything after it is skipped. Use
  it for conditions that make it unsafe to continue.

```{r messaging-warnings, eval=FALSE}
warnings()   # show all recent warnings generated in this session
```

---

# File Handling Functions

Clinical work involves lots of files (data sets, outputs, logs). R can inspect,
create, and manage them without leaving the console.

## Listing files: `dir()` and `list.files()`

`dir()` and `list.files()` are the same function under two names. They list the
contents of a directory.

```{r files-list, eval=FALSE}
dir()                                            # everything in the working directory
dir(path = ".", pattern = "\\.R$", full.names = FALSE)  # only .R files, names only
list.files()                                     # identical to dir()
```

Two arguments you'll use constantly:

- `pattern` — a **regular expression** filter. `"\\.R$"` means "ends with `.R`".
  The `\\.` is an escaped dot (a literal period), and `$` anchors it to the end.
- `full.names` — `TRUE` returns full paths (good for feeding into read
  functions); `FALSE` returns bare file names (good for display).

## Creating, checking, and deleting directories

```{r files-dirs, eval=FALSE}
dir.create("comparison_results")        # make a new folder
dir.exists("comparison_results")        # TRUE if it now exists
unlink("comparison_results", recursive = TRUE)  # delete folder AND its contents
```

`recursive = TRUE` is required to remove a folder that contains files. **Be
careful** — `unlink()` deletes permanently, with no recycle bin.

## Reading many SAS files at once

This is the most powerful pattern in the script. The goal: read *every*
`.sas7bdat` file in a folder into a single named list.

Step 1 — get the full paths of all SAS files:

```{r files-sasfiles, eval=FALSE}
sas_paths <- list.files(
  path      = "~/R_info/sas2R_s/data/sdtm",
  pattern   = "\\.sas7bdat$",   # only SAS transport files
  full.names = TRUE             # full paths, so read_sas can find them
)
sas_paths
```

Step 2 — read them all. `lapply()` applies a function to every element of a
vector and returns a list. Here it reads each path with `haven::read_sas()`:

```{r files-lapply, eval=FALSE}
library(haven)

# Modern pipe style (|>): take the paths, then read each one
sdtm_lst <- list.files("~/R_info/sas2R_s/data/sdtm",
                       pattern = "\\.sas7bdat$", full.names = TRUE) |>
  lapply(haven::read_sas)

# Exactly equivalent, written without the pipe:
sdtm_lst <- lapply(
  list.files("~/R_info/sas2R_s/data/sdtm",
             pattern = "\\.sas7bdat$", full.names = TRUE),
  haven::read_sas
)
```

Step 3 — give the list meaningful names. Right now `sdtm_lst` is indexed by
number; we want to call data sets by name (e.g. `sdtm_lst$ae`). We take the file
names, strip the folder with `basename()`, and strip the extension with
`gsub()`:

```{r files-names, eval=FALSE}
names(sdtm_lst) <- gsub(
  "\\.sas7bdat$", "",                       # remove the extension
  basename(list.files("~/R_info/sas2R_s/data/sdtm",
                      pattern = "\\.sas7bdat$", full.names = FALSE))
)
```

Step 4 — access data sets in the list. There are several ways to index:

```{r files-access, eval=FALSE}
sdtm_lst[[1]]              # first data set (by position)
sdtm_lst$ae                # the "ae" data set (by name)
sdtm_lst$ae[1, 1]          # row 1, column 1 of ae
sdtm_lst$ae[, "USUBJID"]   # the whole USUBJID column of ae
```

Quick reminder on brackets (from the script's own note):

- `()` **calls** a function — `mean(x)`.
- `[]` **subsets** a vector, list, or data frame — `x[1]`, `df[, "AGE"]`.
- `[[]]` extracts a **single element** from a list — `sdtm_lst[["ae"]]`.
- `{}` **groups** several expressions, e.g. a function body.

## Inspecting and copying files

```{r files-info, eval=FALSE}
# Details (size, modified time, permissions) for every SAS file
file.info(list.files("~/R_info/sas2R_s/data/sdtm",
                     pattern = "\\.sas7bdat$", full.names = TRUE))

# Back up all SAS files to another folder
file.copy(
  from = list.files("~/R_info/sas2R_s/data/sdtm",
                    pattern = "\\.sas7bdat$", full.names = TRUE),
  to   = "~/R_info/sas2R_s/data/sdtm_backup"
)

file.exists("~/R_info/sas2R_s/data/sdtm_backup")   # confirm the backup exists
```

---

# Writing Your Own Functions

Once you repeat the same steps twice, wrap them in a function.

## The basic syntax

```{r functions-syntax}
# name <- function(arguments) { body }
test <- function(x) {
  y <- x + 1
  print(y)
}

test(2)   # call the function with the argument 2
```

The four parts of a function:

1. **Name** — how you call it later (`test`).
2. **Arguments** — inputs in parentheses (`x`).
3. **Body** — the code in `{ }` that does the work.
4. **Return value** — what comes back. R returns the last evaluated expression;
   you can also be explicit with `return()`.

## Documenting a function (roxygen2)

Special comments starting with `#'` are **roxygen2** documentation. When the
function lives in a package, roxygen turns these into a proper help page (the
thing you see when you type `?test`).

```{r functions-doc, eval=FALSE}
#' Test Function
#'
#' This function takes a numeric input and adds 1 to it.
#'
#' @param x A numeric value.
#' @return The input value incremented by 1.
#' @examples
#' test(2)  # returns 3
#' test(5)  # returns 6
#' @export
test <- function(x) {
  y <- x + 1
  print(y)
}
```

The tags mean: `@param` documents each argument, `@return` describes the output,
`@examples` gives runnable examples, and `@export` makes the function visible to
users of your package.

## Three kinds of functions

- **Built-in** — ship with base R: `mean()`, `sum()`, `length()`.
- **User-defined** — ones you write with `function`, like `test()` above.
- **Package** — come from installed packages: `dplyr::filter()`,
  `ggplot2::ggplot()`.

---

# Exploring Existing Packages and Functions

Before writing new code, learn what already exists. R is very introspectable —
you can ask it about itself.

## What is CRAN?

**CRAN** (the Comprehensive R Archive Network) is the central, official
repository of R packages. When you run `install.packages("dplyr")`, R downloads
it from CRAN. It stores the packages, their source code, and documentation.

## Discovering a package's contents

```{r explore, eval=FALSE}
library(help = "dplyr")     # opens a list of dplyr's functions and data sets
ls()                        # list objects in YOUR current environment
library(dplyr)              # load the package so its functions are available
ls("package:dplyr")         # list every function dplyr exposes
objects("package:dplyr")    # same as ls() — list the package's objects
```

## Looking inside a specific function

```{r explore-func, eval=FALSE}
args(dplyr::filter)   # what arguments does filter() accept?
body(dplyr::filter)   # show the actual source code of filter()
```

`args()` and `body()` are great learning tools: `args()` shows the "signature"
(the inputs), and `body()` reveals how the function actually works.

---

# Git Integration with RStudio

**Git** is version control — it saves snapshots ("commits") of your project so
you can track history and collaborate. **GitHub** is a website that hosts Git
repositories online.

## The everyday cycle (in RStudio's Git pane)

1. **Initialize** a Git repository in your project directory.
2. **Create** a matching repository on GitHub and copy its URL.
3. In RStudio's **Git** tab, click **Commit**, stage your changed files, and
   write a short message describing the change.
4. Click **Push** to upload your commits to GitHub.
5. Click **Pull** to bring down changes made by others.

## One-time configuration

Tell Git who you are (this stamps every commit with your name):

```{r git-config, eval=FALSE}
library(usethis)
use_git_config(user.name = "My Username", user.email = "my_email@address.com")
```

You can also authenticate to GitHub. Modern GitHub uses either an SSH key
(Tools → Global Options → Git/SVN → Create RSA Key) or a **Personal Access
Token (PAT)**:

```{r git-auth, eval=FALSE}
usethis::edit_git_config()        # open the git config file to inspect/edit
usethis::create_github_token()    # opens GitHub to mint a PAT
library(gitcreds)
gitcreds::gitcreds_set()          # paste the PAT to store it securely

# lower-level alternative to set config directly:
gert::git_config_set("user.email", "my_email@address.com")
```

## Wiring a project to Git and GitHub from R

`usethis` automates the whole setup:

```{r git-usethis, eval=FALSE}
usethis::use_git()      # initialise the repo + make the first commit
usethis::use_r("my_new_script")  # create R/my_new_script.R in a package
usethis::use_github()   # create the GitHub repo and connect it to your project
```

---

# Building Your First Package

A **package** is the standard way to bundle functions, documentation, and tests
so you (and others) can reuse them reliably.

## Install the development toolkit

```{r pkg-install, eval=FALSE}
install.packages(c("devtools", "roxygen2", "usethis", "testthat"))
```

- **usethis** — automates project/package setup tasks.
- **roxygen2** — turns `#'` comments into help pages.
- **devtools** — the build/test/check/install workflow.
- **testthat** — the unit-testing framework.

## Package-level documentation

Just as functions get roxygen comments, so does the whole package. A special
block describes the package itself:

```{r pkg-doc, eval=FALSE}
#' My First R Package
#'
#' This package contains functions for basic arithmetic operations.
#'
#' @docType package
#' @name mypackage
#' @keywords package
#' @examples
#' test(2)
#' test(5)
"_PACKAGE"
```

## The development loop (recap)

```{r pkg-loop, eval=FALSE}
usethis::create_package("~/projects/mypackage")  # scaffold the package
usethis::use_r("test")        # create R/test.R and paste your function there
usethis::use_testthat()       # set up the tests/ folder
usethis::use_test("test")     # create a matching test file

devtools::document()   # build help pages + NAMESPACE from roxygen comments
devtools::load_all()   # load the package for interactive testing
devtools::test()       # run all tests
devtools::check()      # full R CMD check — the quality gate
devtools::install()    # install it into your R library
```

---

# rlang: Tidy Evaluation for Beginners

The **rlang** package powers the "magic" in the tidyverse — the way `dplyr` lets
you type a bare column name like `filter(df, AGE > 40)` without quotes. rlang
gives you tools to **capture code**, **manipulate expressions**, and **write your
own tidyverse-style functions**.

```{r rlang-load}
library(rlang)
library(dplyr)
```

## Why it exists

When you write `filter(adsl_dev, AGE > 40)`, there is no object called `AGE` in
your workspace — yet it works. dplyr **captures** the expression `AGE > 40` and
evaluates it *inside* the data frame. This is **non-standard evaluation (NSE)**,
and rlang is the toolkit that makes it possible.

## Core rlang functions

```{r rlang-core}
# expr(): capture an expression WITHOUT running it
e <- expr(AGE + 1)
e
class(e)          # a "call" — a stored piece of code

# sym(): turn a string into a symbol (a bare name)
s <- sym("AGE")
s

# eval_tidy(): evaluate captured code inside a data context
eval_tidy(expr(AGE * 2), data = adsl_dev)

# as_label(): turn an expression/quosure back into a readable string
as_label(e)
```

## Writing a function that takes a column name

The most useful beginner skill: a wrapper around dplyr that accepts a column
name. The naive version fails because dplyr takes the literal word `col`:

```{r rlang-broken, error=TRUE}
avg_broken <- function(data, col) {
  summarise(data, avg = mean(col))   # looks for a column literally named "col"
}
avg_broken(adsl_dev, AGE)
```

The fix is `{{ }}` ("curly-curly"), which says "use whatever the caller typed":

```{r rlang-curly}
avg <- function(data, col) {
  summarise(data, avg = mean({{ col }}))
}
avg(adsl_dev, AGE)   # works: AGE flows through correctly
```

Under the hood, `{{ col }}` is shorthand for the classic two-step rlang idiom the
script lists — `enquo()` to **capture**, then `!!` ("bang-bang") to **unquote**:

```{r rlang-enquo}
avg_classic <- function(data, col) {
  col <- enquo(col)                    # capture what the user typed
  summarise(data, avg = mean(!!col))   # inject it back into the dplyr call
}
avg_classic(adsl_dev, AGE)
```

If the column name arrives as a **string** instead, use the `.data` pronoun:

```{r rlang-string}
avg_str <- function(data, col) {
  summarise(data, avg = mean(.data[[col]]))
}
avg_str(adsl_dev, "AGE")   # note the quotes
```

## The rlang "is_*" predicate family

The script lists many `is_*()` helpers. These are **type-checking predicates** —
each returns a single `TRUE`/`FALSE`. They're handy for input validation because
they're consistent and read clearly.

```{r rlang-predicates}
is_symbol(sym("AGE"))          # TRUE — it's a symbol
is_call(expr(mean(AGE)))       # TRUE — it's a function call
is_quosure(quo(AGE))           # TRUE — quo() makes a quosure
is_atomic(1:5)                 # TRUE — a plain atomic vector
is_list(list(1, 2))            # TRUE
is_function(mean)              # TRUE
is_null(NULL)                  # TRUE
is_empty(list())               # TRUE — length zero
is_named(c(a = 1, b = 2))      # TRUE — the vector has names
```

A quick tour of what the others check (from the script):

- `is_expression()` — an expression object.
- `is_formula()` — a formula like `y ~ x`.
- `is_vector()` / `is_data_frame()` / `is_tibble()` — container types.
- `is_missing()` — whether a function argument was left out.
- `is_finite()` / `is_infinite()` / `is_nan()` — numeric edge cases
  (`Inf`, `-Inf`, `NaN`).

These behave more predictably than some base R equivalents, which is why
tidyverse code favours them.

---

# Putting It All Together

A realistic mini-workflow that uses several ideas from this tutorial: compare two
data sets, log the outcome with the messaging functions, and save the report.

```{r together, eval=FALSE}
library(diffdf)

compare_and_log <- function(dev, qc, out_dir = "comparison_results") {
  # 1. make an output folder if needed
  if (!dir.exists(out_dir)) dir.create(out_dir)

  # 2. quick gate
  if (identical(dev, qc)) {
    message("Data sets are identical — nothing to report.")
    return(invisible(NULL))
  }

  # 3. detailed comparison
  result <- diffdf(dev, qc)

  # 4. communicate + save
  warning("Differences found — review before proceeding.")
  saveRDS(result, file.path(out_dir, "diff_result.rds"))
  message("Comparison complete. Report saved to ", out_dir)

  invisible(result)
}

compare_and_log(adsl_dev, adsl_qc)
```

---

# Summary

From the `package_dev.R` script you have now met:

- **Comparison** — `identical()`, `diffdf()`, and `arsenal::comparedf()`.
- **Communication** — `print()`, `message()`, `warning()`, `stop()`, `warnings()`.
- **File handling** — `dir()`, `list.files()`, `dir.create()`, `unlink()`,
  `file.info()`, `file.copy()`, and reading many SAS files with `lapply()`.
- **Functions** — syntax, the four parts, and roxygen2 documentation.
- **Introspection** — `library(help=)`, `ls()`, `args()`, `body()`, and CRAN.
- **Git** — configuration and the commit/push/pull cycle via `usethis`.
- **Packages** — the `usethis` + `devtools` + `roxygen2` build loop.
- **rlang** — `expr()`, `sym()`, `enquo()`, `!!`, `{{ }}`, `.data[[ ]]`,
  `eval_tidy()`, and the `is_*()` predicate family.

The professional habit that ties it together:
**write → document → test → commit → check → push.**

## Further reading

- *R Packages* (2nd ed.) — https://r-pkgs.org
- *Advanced R* — https://adv-r.hadley.nz
- diffdf — https://cran.r-project.org/package=diffdf
- arsenal — https://cran.r-project.org/package=arsenal
- rlang / tidy evaluation — https://rlang.r-lib.org and
  https://dplyr.tidyverse.org/articles/programming.html
- Happy Git with R — https://happygitwithr.com
